# How a Plot Can Mislead {.unnumbered}

## Where this fits

You have a clean, captioned worked example ([viz03](viz03-worked-example.qmd)). This page is the audit lens: six common ways a figure that *looks fine* can quietly mislead a reader. Each one shows up in real research, and each one is something an AI co-pilot will produce if you do not push back.

## Six failure modes

For each failure mode, you will see the problem in one sentence and the fix in one sentence. Studio time in [viz06](viz06-studio-visual-audit.qmd) revisits these on a deliberately flawed plot.

### 1. Truncated y-axis exaggeration

A small difference in means looks huge when the y-axis starts at 27 instead of 0.

- **Symptom:** `coord_cartesian(ylim = c(27.5, 31.5))` on a BMI mean plot, no narrative explanation.
- **Fix:** Start the axis at 0, or at a value justified in the caption. Use `expand_limits(y = 0)` when in doubt.

### 2. Causal title on a descriptive plot

A title that says "X causes Y" is a causal claim. Descriptive summaries cannot support it.

- **Symptom:** "Income level causes BMI to rise across NHANES cycles."
- **Fix:** "Mean BMI by income group across NHANES cycles in the prepared classroom dataset."

<!-- AI-EDIT(2026-06-11): TF-018, WI-035 — needs review -->
::: {.callout-note title="Definition: causal overreach"}
A descriptive chart can show a pattern or association, but it does not prove that one factor caused another. Claiming cause from a descriptive plot is called causal overreach.
:::

### 3. Smoothing across categorical time

`geom_smooth(method = "lm")` draws a line that *implies* a continuous linear trend. NHANES cycles are categorical labels, not a continuous axis.

- **Symptom:** A linear regression line layered across cycle bins.
- **Fix:** Use `geom_line()` plus `geom_point()` to connect descriptive cycle summaries, with no regression overlay.

### 4. Weighted-vs-unweighted ambiguity

If the subtitle says "weighted" but the code does not use a survey weight variable, the reader has been misled.

- **Symptom:** Subtitle: "Weighted BMI trend." Code: `mean(BMI, na.rm = TRUE)` with no weight.
- **Fix:** State plainly that the summary is unweighted, or compute a weighted estimate using the documented weight variable.

### 5. Silent missingness

Rows with missing BMI or income disappear without a caption.

- **Symptom:** `filter(!is.na(BMI), !is.na(IncomeGroup))` deep inside a pipe with no note in the caption.
- **Fix:** Disclose missing-value exclusions in the caption. Better: report missingness counts before the figure (you will do this on [eda03](../week07-eda-ai-audit/eda03-worked-example.qmd) next week).

### 6. Hidden small-cell counts

A group with `n = 7` plots the same way as a group with `n = 700`, but the reader cannot tell them apart.

- **Symptom:** A small income group plotted as a line with no indication of sparse cells.
- **Fix:** Filter out groups below a minimum count, *or* label them visibly with `n` annotations, *or* report the counts beside the figure.

<!-- AI-EDIT(2026-06-11): TF-018, WI-034 — needs review -->
::: {.callout-note title="Definition: hidden small cell count"}
A hidden small cell count is a group with very few observations that may be hidden inside a chart or summary. Do not make strong claims from these groups.
:::

<!-- AI-EDIT(2026-06-11): TF-019, MIT-065, SS-033 — visible before/after plot pair, generated at render time — needs review -->
## Seeing it: a before/after pair

Reading about failure modes is one thing; seeing them side by side is another. The two figures below use the same summary as the worked example on [viz03](viz03-worked-example.qmd). The first plot combines failure modes #1 and #2. The second plot is the honest version. (This pair is *not* the studio plot — the flawed plot you will audit in [viz06](viz06-studio-visual-audit.qmd) has its own problems for you to find.)

```{r}
#| label: viz04-prepare-summary
#| message: false
#| warning: false
library(readr)
library(dplyr)
library(ggplot2)

candidate_paths <- c(
  "examples/nhanes-equity/data/nhanes_equity_v6.csv",
  "../../examples/nhanes-equity/data/nhanes_equity_v6.csv"
)
data_path <- candidate_paths[file.exists(candidate_paths)][1]
nhanes <- read_csv(data_path, show_col_types = FALSE)

bmi_by_income_cycle <- nhanes |>
  filter(!is.na(BMI), !is.na(IncomeGroup), !is.na(Cycle), Age >= 20, Age <= 80) |>
  group_by(Cycle, IncomeGroup) |>
  summarise(n = n(), mean_bmi = mean(BMI), .groups = "drop") |>
  filter(n >= 30)
```

**Before — truncated axis and causal title.** The y-axis starts near the data, so small gaps look dramatic, and the title makes a claim the descriptive summary cannot support:

```{r}
#| label: viz04-before-misleading
#| fig-alt: "Misleading line plot of mean BMI by income group with a truncated y-axis from 27.5 to 31.5 and the causal title 'Income level causes BMI to rise across NHANES cycles'."
ggplot(
  bmi_by_income_cycle,
  aes(x = Cycle, y = mean_bmi, color = IncomeGroup, group = IncomeGroup)
) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2.5) +
  scale_color_brewer(palette = "Dark2") +
  coord_cartesian(ylim = c(27.5, 31.5)) +
  labs(
    title = "Income level causes BMI to rise across NHANES cycles",
    x = "NHANES cycle",
    y = "Mean BMI",
    color = "Income group"
  ) +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "bottom")
```

**After — honest scale and descriptive title.** Same data, same geometry. The axis starts at 0, the title describes instead of claiming cause, and the caption discloses the exclusions:

```{r}
#| label: viz04-after-corrected
#| fig-alt: "Corrected line plot of mean BMI by income group across NHANES cycles with the y-axis starting at 0, a descriptive title, and a caption disclosing exclusions and the unweighted descriptive limitation."
ggplot(
  bmi_by_income_cycle,
  aes(x = Cycle, y = mean_bmi, color = IncomeGroup, group = IncomeGroup)
) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2.5) +
  scale_color_brewer(palette = "Dark2") +
  expand_limits(y = 0) +
  labs(
    title = "Mean BMI by income group across NHANES cycles",
    subtitle = "Adults age 20-80 in the prepared class dataset; unweighted descriptive summary",
    x = "NHANES cycle",
    y = "Mean BMI",
    color = "Income group",
    caption = "Records with missing BMI, income, or cycle excluded; groups under n = 30 removed.\nDescriptive only, not a causal or population estimate."
  ) +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "bottom")
```

Notice what the fix did *not* change: the data, the geometry, and the colors are identical. Honest visualization is mostly about scale, title, and disclosure.

## Why six and not sixty

You will see other failure modes in the wild — duplicated colors, default rainbow palettes, dual axes, missing legends. The six above account for most of the issues you will catch when auditing an AI-drafted plot for descriptive health data work in this course. Master these first.

## A quick audit habit

When you finish a plot, ask three questions out loud:

1. Could a reader make a causal claim from this title or caption?
2. Could a small difference look bigger than it is because of the axis?
3. Is missingness, exclusion, or small-cell count visible to the reader?

If any answer is "yes" or "I don't know," revise the figure or the caption before moving on.

## Where this goes next

[viz05](viz05-ai-prompts-and-audit.qmd) shows how to write AI prompts that produce plots you can defend, and how to audit what the AI gives you.
