# Worked Example: Descriptive BMI Visualization {.unnumbered}

## Where this fits

You learned the grammar of a ggplot and made one minimal scatter on the [previous page](viz02-grammar-and-first-plot.qmd). This page is the first end-to-end visualization of the term: load the cached NHANES Health Equity CSV, prepare a small descriptive summary, render one labeled figure, and write a caption that does not overclaim. We will return to this same figure in [viz04](viz04-misleading-design.qmd) to see how easy it is to spoil with a few "improvements."

## Goal

This worked example uses the NHANES Health Equity CSV snapshot to build one descriptive visualization. The plot summarizes mean BMI by income group across NHANES cycle. It is **unweighted** and should be interpreted as a classroom visualization exercise, not a population estimate.

<!-- AI-EDIT(2026-06-11): TF-018, MIT-066, WI-033 — needs review -->
::: {.callout-note title="Definition: weighted vs unweighted"}
**Weighted** means the estimate accounts for survey design. **Unweighted** means it only summarizes the observed classroom/cache data. Use unweighted claims only as descriptive classroom practice.
:::

```{r}
#| label: load-nhanes-equity-visualization
#| message: false
#| warning: false
library(readr)
library(dplyr)
library(ggplot2)

candidate_paths <- c(
  "examples/nhanes-equity/data/nhanes_equity_v6.csv",
  "../../examples/nhanes-equity/data/nhanes_equity_v6.csv"
)

data_path <- candidate_paths[file.exists(candidate_paths)][1]
if (is.na(data_path)) {
  stop("Could not find examples/nhanes-equity/data/nhanes_equity_v6.csv")
}

nhanes <- read_csv(data_path, show_col_types = FALSE)
```

## Prepare a Descriptive Summary

Keep the summary intentionally simple. We are describing the prepared class dataset, not estimating a national trend.

```{r}
#| label: summarize-bmi-by-income-cycle
bmi_by_income_cycle <- nhanes |>
  filter(
    !is.na(BMI),
    !is.na(IncomeGroup),
    !is.na(Cycle),
    Age >= 20,
    Age <= 80
  ) |>
  group_by(Cycle, IncomeGroup) |>
  summarise(
    n = n(),
    mean_bmi = mean(BMI),
    .groups = "drop"
  ) |>
  filter(n >= 30)

bmi_by_income_cycle
```

<!-- AI-EDIT(2026-06-11): TF-018, MIT-066, WI-034 — needs review -->
::: {.callout-note title="Definition: hidden small cell count"}
A hidden small cell count is a group with very few observations that may be hidden inside a chart or summary. Do not make strong claims from these groups. That is why the code above keeps only groups with `n >= 30`.
:::

## Plot

<!-- AI-EDIT(2026-06-11): MIT-063 — thicker lines + Dark2 palette so the bottom line reads clearly — needs review -->
<!-- AI-EDIT(2026-06-22): MIT-063 / REVIEW-row154 — the long in-plot caption did not wrap and got clipped at the bottom; moved the descriptive note to a Quarto fig-cap (which wraps below the figure), shortened the in-plot caption, added fig-height/fig-width and a plot.caption bottom margin so nothing is cut off -->
```{r}
#| label: plot-bmi-by-income-cycle
#| fig-height: 5.5
#| fig-width: 7
#| fig-cap: "Mean BMI is summarized descriptively from the cached class dataset. Values are unweighted, exclude missing BMI/income/cycle values, and should not be interpreted as causal effects or population estimates."
bmi_plot <- ggplot(
  bmi_by_income_cycle,
  aes(x = Cycle, y = mean_bmi, color = IncomeGroup, group = IncomeGroup)
) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2.5) +
  scale_color_brewer(palette = "Dark2") +
  labs(
    title = "Mean BMI by income group across NHANES cycle",
    subtitle = "Adults age 20-80 in the prepared class dataset; unweighted descriptive summary",
    x = "NHANES cycle",
    y = "Mean BMI",
    color = "Income group",
    caption = "Unweighted descriptive summary; not a population estimate."
  ) +
  theme_minimal(base_size = 12) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "bottom",
    plot.caption = element_text(margin = margin(t = 8))
  )

bmi_plot
```

## Optional Plotly Check

<!-- AI-EDIT(2026-06-11): MIT-064 — needs review -->
{{< include /_snippets/demo-only.qmd >}}

`plotly` is an R package that turns a static ggplot into an interactive web graphic — you can hover over points to see values and zoom into regions. <!-- AI-EDIT(2026-06-11): MIT-064 — needs review -->

The core assignment uses the static `ggplot2` figure above. If `plotly` is installed, you can preview the same plot interactively while checking labels and tooltips. This extension is optional because static Quarto output is the supported submission path.

```{r}
#| label: optional-plotly-preview
#| eval: false
# plotly::ggplotly(bmi_plot)
```

## Caption Template

Use this structure when revising your own plot:

> This figure shows [measure] by [group] across [time/category] using the prepared NHANES Health Equity classroom dataset. The summary is descriptive and unweighted. Records with missing [variables] were excluded. The plot supports comparison of patterns, not causal claims or population-level estimates.

## Mini Visual Audit Checklist

- Does the title describe what is plotted without overstating the claim?
- Is it clear whether the statistic is weighted or unweighted?
- Are axes, groups, and units labeled plainly?
- Are missing values or exclusions disclosed?
- Is the scale chosen to support honest comparison rather than exaggeration?
- Does the caption state the limitation that this is descriptive, not causal?

## Where this goes next

Hold onto this figure. The [next page](viz04-misleading-design.qmd) walks through six common ways a plot like this can mislead a reader, and [viz06](viz06-studio-visual-audit.qmd) is the in-class studio where you apply the audit to a deliberately flawed version of it.
