# Grammar of a ggplot, and Your First Plot {.unnumbered}

## Where this fits

[viz01](viz01-overview-and-goals.qmd) framed visualization as a claim about evidence. This page teaches the five moving parts of any `ggplot2` figure and walks you through your first plot on the NHANES Health Equity CSV. Everything builds from these five parts.

## The five moving parts

Every ggplot is built from the same five ingredients. You will see all five in the worked example on the next page, so the names matter.

| Part | What it answers | A beginner-friendly analogy |
|---|---|---|
| **Data** | What table are we plotting? | The contents of one box |
| **Mapping (`aes`)** | Which variable goes where (x, y, color, group)? | Telling R which column to draw on which axis |
| **Geometry (`geom_*`)** | What kind of mark (point, line, bar)? | The shape of the ink on the page |
| **Scale and coordinates** | How are values stretched along an axis? | The ruler under the ink |
| **Labels and theme** | What does the reader see in plain text? | The caption around the picture |

If a reader cannot answer "which statistic? across what groups? with what limitation?" from your figure, one of these five parts is doing too little.

## Setup

Open your Codespace and start a new code chunk in a `.qmd` file. Run:

<!-- AI-EDIT(2026-06-23): tidyverse-default — consolidated core library() calls into library(tidyverse) -->
```{r}
#| eval: false
library(tidyverse)
```

::: {.callout-warning}
`library(tidyverse)` is required every session. If you see "could not find function `ggplot`", run the `library()` line first.
:::

## Load the cached NHANES CSV

This is the same file you will use for the worked example and your A5 deliverable. The candidate-path pattern below works whether your `.qmd` is at the repository root or under `weeks/week06-visualization/`.

```{r}
#| eval: false
candidate_paths <- c(
  "examples/nhanes-equity/data/nhanes_equity_v6.csv",
  "../../examples/nhanes-equity/data/nhanes_equity_v6.csv"
)
data_path <- candidate_paths[file.exists(candidate_paths)][1]
nhanes <- read_csv(data_path, show_col_types = FALSE)

glimpse(nhanes)
```

`glimpse()` shows the column names and types. Read them carefully. Misspelling a column name (e.g., `bmi` when the data has `BMI`) is the #1 source of broken ggplots in this course.

<!-- AI-EDIT(2026-06-22): tester-note-conversion — converted leaked "Tester note" callout to hidden HTML comment (tester instruction, not student content)
Tester note — Capture: a screenshot of the Codespaces Quarto preview after `glimpse(nhanes)` runs, showing the column names. New students rely on this to spot case-sensitive column names like `BMI` and `IncomeGroup`.
-->

## Your first plot

Here is the smallest meaningful ggplot you can build from this dataset: a scatter of one numeric variable against age, with no grouping and no transformation. Try it as is, then change `Weight` to `BMI` or `Height` and re-render.

```{r}
#| eval: false
ggplot(nhanes, aes(x = Age, y = Weight)) +
  geom_point(alpha = 0.2) +
  labs(
    title = "Weight by age, NHANES Health Equity classroom dataset",
    subtitle = "One point per record; unweighted, descriptive only",
    x = "Age (years)",
    y = "Weight (kg)"
  ) +
  theme_minimal()
```

Map the five parts onto this code so the grammar feels concrete:

- **Data:** `nhanes`
- **Mapping:** `aes(x = Age, y = Weight)`
- **Geometry:** `geom_point(alpha = 0.2)` (translucent points so overlap is visible)
- **Scale and coordinates:** the default linear axes (you have not changed them)
- **Labels and theme:** `labs(...) + theme_minimal()`

<!-- AI-EDIT(2026-06-22): tester-note-conversion — converted leaked "Tester note" callout to hidden HTML comment (tester instruction, not student content)
Tester note — Capture: a screenshot of the rendered scatter plot. Confirm the axis labels read "Age (years)" and "Weight (kg)" so students can compare against their own first attempt.
-->

## Three things to check on every plot

After you run any ggplot, ask:

1. **Column names match the data.** Did `glimpse()` confirm `BMI` is uppercase, not `bmi`?
2. **Axes tell a clear story.** Are the units readable? Is the range honest?
3. **Title and subtitle match what is plotted.** Does the title overstate ("BMI rises with age across Canada")? If yes, soften it ("Weight by age in the prepared classroom dataset").

## Where this goes next

[viz03](viz03-worked-example.qmd) is the first end-to-end worked example: load the CSV, prepare a small descriptive summary, render a labeled figure, and write a caption.
