# Worked Example: Descriptive BMI Visualization {.unnumbered}

## Goal

This worked example uses the NHANES Health Equity CSV snapshot to build one descriptive visualization. The plot summarizes mean BMI by income group across NHANES cycle. It is **unweighted** and should be interpreted as a classroom visualization exercise, not a population estimate.

```{r}
#| label: load-nhanes-equity-visualization
#| message: false
#| warning: false
library(readr)
library(dplyr)
library(ggplot2)

candidate_paths <- c(
  "examples/nhanes-equity/data/nhanes_equity_v6.csv",
  "../../examples/nhanes-equity/data/nhanes_equity_v6.csv"
)

data_path <- candidate_paths[file.exists(candidate_paths)][1]
if (is.na(data_path)) {
  stop("Could not find examples/nhanes-equity/data/nhanes_equity_v6.csv")
}

nhanes <- read_csv(data_path, show_col_types = FALSE)
```

## Prepare a Descriptive Summary

Keep the summary intentionally simple. We are describing the prepared class dataset, not estimating a national trend.

```{r}
#| label: summarize-bmi-by-income-cycle
bmi_by_income_cycle <- nhanes |>
  filter(
    !is.na(BMI),
    !is.na(IncomeGroup),
    !is.na(Cycle),
    Age >= 20,
    Age <= 80
  ) |>
  group_by(Cycle, IncomeGroup) |>
  summarise(
    n = n(),
    mean_bmi = mean(BMI),
    .groups = "drop"
  ) |>
  filter(n >= 30)

bmi_by_income_cycle
```

## Plot

```{r}
#| label: plot-bmi-by-income-cycle
ggplot(
  bmi_by_income_cycle,
  aes(x = Cycle, y = mean_bmi, color = IncomeGroup, group = IncomeGroup)
) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 2) +
  labs(
    title = "Mean BMI by income group across NHANES cycle",
    subtitle = "Adults age 20-80 in the prepared class dataset; unweighted descriptive summary",
    x = "NHANES cycle",
    y = "Mean BMI",
    color = "Income group",
    caption = "Caption template: Mean BMI is summarized descriptively from the cached class dataset. Values are unweighted, exclude missing BMI/income/cycle values, and should not be interpreted as causal effects or population estimates."
  ) +
  theme_minimal(base_size = 12) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "bottom"
  )
```

## Caption Template

Use this structure when revising your own plot:

> This figure shows [measure] by [group] across [time/category] using the prepared NHANES Health Equity classroom dataset. The summary is descriptive and unweighted. Records with missing [variables] were excluded. The plot supports comparison of patterns, not causal claims or population-level estimates.

## Mini Visual Audit Checklist

- Does the title describe what is plotted without overstating the claim?
- Is it clear whether the statistic is weighted or unweighted?
- Are axes, groups, and units labeled plainly?
- Are missing values or exclusions disclosed?
- Is the scale chosen to support honest comparison rather than exaggeration?
- Does the caption state the limitation that this is descriptive, not causal?
