# What Each Statistic Claims (and Does Not) {.unnumbered}

## Where this fits

You produced a Table 1 on the [previous page](eda03-worked-example.qmd). Before you write an interpretation of it, pause and ask: what does each cell in that table actually *claim*? This page is a short tour through the five most common descriptive statistics you will see — what each one says, what it leaves out, and the most common ways beginners (and AI tools) mislabel them.

## The five common statistics

| Statistic | What it answers | What it does not answer |
|---|---|---|
| **N (count)** | How many records are in this group? | Whether the group is representative of any population |
| **Mean** | What is the average value? | Whether the distribution is symmetric, skewed, or has outliers |
| **SD (standard deviation)** | How spread out are values around the mean? | Whether the spread is symmetric or driven by a tail |
| **Median** | What value falls in the middle? | How extreme the tails are |
| **IQR (interquartile range)** | What is the spread of the middle 50%? | The tails outside the middle 50% |

A Table 1 typically reports **N** plus either **mean (SD)** or **median (IQR)** depending on the variable.

## When to prefer mean(SD) vs median(IQR)

Use **mean (SD)** when the variable is roughly symmetric and well-behaved. Age in the adult cohort is a reasonable example.

Use **median (IQR)** when the variable is skewed or has outliers. Healthcare costs, hospital length of stay, and household income are usual suspects. The mean of a skewed variable is pulled by the long tail; the median is not.

If you are unsure, plot a histogram. If you see a long tail on one side, prefer median (IQR).

::: {.callout-tip}
For this course's NHANES classroom Table 1, BMI and age are commonly reported as mean (SD). That is a reasonable convention for adult anthropometric variables. State the choice and the reason in your methods note.
:::

## What "N" alone does not tell you

A cell that says `N = 47` is informative only if the reader knows:

- which cohort the 47 came from (cohort definition, [eda02](eda02-cohort-and-missingness.qmd));
- how many of those 47 were excluded for missing values (missingness table, also [eda02](eda02-cohort-and-missingness.qmd));
- what the comparison group sizes are (the rest of the Table 1).

If any of those three is invisible, the `N` is not yet defensible.

## Three common beginner mistakes

These are the mistakes that show up most often in AI-drafted Table 1 code. The [studio](eda06-studio-planted-errors.qmd) revisits them on a deliberately flawed starter.

### 1. Label says "mean" but code computes `median()`

```{r}
#| eval: false
# WRONG: label and code disagree
summarise(`Mean BMI` = median(BMI, na.rm = TRUE))
```

Either change the label to `Median BMI` or change the function to `mean()`. Decide which one you want — do not let the AI choose silently.

### 2. SD reported without `na.rm = TRUE`

```{r}
#| eval: false
# WRONG: returns NA whenever any BMI value is missing
summarise(sd_bmi = sd(BMI))
```

If a single `NA` is present, `sd(BMI)` returns `NA`. Use `na.rm = TRUE` and report the missingness count separately (see [eda02](eda02-cohort-and-missingness.qmd)).

### 3. Percent without a denominator

```{r}
#| eval: false
# WRONG: a "60% with low income" cell with no N
summarise(percent_low_income = mean(IncomeGroup == "Low Income (<1.3)") * 100)
```

A percent is only meaningful if the reader can see the denominator. Always include the N in the same row or in an adjacent cell.

## Small-cell suppression: when not to report a number

If a stratified cell has very few records (say, fewer than 10), publishing the exact statistic can be misleading or, in some health data settings, a privacy risk. The convention is to either:

- combine the small group with a neighbor and label it clearly, or
- replace the cell with a "<10" marker.

Your classroom Table 1 with `n >= 30` per group (as in the [worked example](eda03-worked-example.qmd)) is already cautious. The principle still applies: do not interpret cells where N is too small to support the claim.

## The audit habit for descriptive statistics

After you build a Table 1, look at each column heading and each row:

1. Does the column heading describe the statistic *actually* computed (mean, median, count, percent)?
2. Is the denominator visible for every percent?
3. Are missing values either filtered out (with the count reported elsewhere) or handled with `na.rm = TRUE`?
4. Are small cells either combined, marked, or excluded?

If any answer is "no", revise the table or its labels before you write the methods note.

## Where this goes next

[eda05](eda05-ai-audit-categories.qmd) introduces the four AI-audit categories that organize every check you have seen so far this week, ready for the [studio](eda06-studio-planted-errors.qmd) on the planted-error starter.
