# Report Template: NHANES Equity Summary {.unnumbered}

## Standalone YAML

If you copy this template into a project report outside the book, start the file with this YAML header:

```yaml
---
title: "NHANES Equity Summary"
format: html
bibliography: ../../references.bib
---
```

## Purpose

This runnable template shows how to pair a dashboard with a short reproducible report. The dashboard supports exploration; the report fixes one question, documents the data, and cites the source.

## Question

How does mean BMI vary by income group among adults age 20-80 in the prepared NHANES Health Equity classroom dataset?

## Data Source

The report uses the cached CSV snapshot at `examples/nhanes-equity/data/nhanes_equity_v6.csv`. The source data are derived from CDC/NCHS NHANES public-use files [@cdc-nhanes].

```{r}
#| message: false
#| warning: false
library(readr)
library(dplyr)
library(ggplot2)
library(knitr)

candidate_paths <- c(
  "examples/nhanes-equity/data/nhanes_equity_v6.csv",
  "../../examples/nhanes-equity/data/nhanes_equity_v6.csv"
)

data_path <- candidate_paths[file.exists(candidate_paths)][1]
if (is.na(data_path)) {
  stop("Could not find examples/nhanes-equity/data/nhanes_equity_v6.csv")
}

nhanes <- read_csv(data_path, show_col_types = FALSE)
```

## Methods

We restricted the classroom dataset to adults age 20-80 and summarized BMI by income group. The summaries are unweighted and descriptive. They should not be interpreted as causal effects or national population estimates.

```{r}
analysis_df <- nhanes |>
  filter(Age >= 20, Age <= 80)

missingness <- analysis_df |>
  summarise(
    rows = n(),
    missing_bmi = sum(is.na(BMI)),
    missing_income = sum(is.na(IncomeGroup)),
    missing_age = sum(is.na(Age))
  )

kable(missingness, caption = "Missingness check for key report variables.")
```

## Results

```{r}
income_summary <- analysis_df |>
  filter(!is.na(BMI), !is.na(IncomeGroup)) |>
  group_by(IncomeGroup) |>
  summarise(
    n = n(),
    mean_bmi = mean(BMI),
    sd_bmi = sd(BMI),
    .groups = "drop"
  ) |>
  mutate(
    `BMI, mean (SD)` = sprintf("%.1f (%.1f)", mean_bmi, sd_bmi)
  ) |>
  select(IncomeGroup, n, `BMI, mean (SD)`)

kable(income_summary, caption = "Descriptive BMI summary by income group.")
```

```{r}
#| fig-cap: "Mean BMI by income group among adults age 20-80 in the prepared NHANES Health Equity classroom dataset. Values are unweighted and descriptive."
plot_df <- analysis_df |>
  filter(!is.na(BMI), !is.na(IncomeGroup)) |>
  group_by(IncomeGroup) |>
  summarise(
    n = n(),
    mean_bmi = mean(BMI),
    .groups = "drop"
  )

ggplot(plot_df, aes(x = IncomeGroup, y = mean_bmi)) +
  geom_col(fill = "#2C7FB8") +
  geom_text(aes(label = paste0("n=", n)), vjust = -0.4, size = 3.5) +
  labs(
    title = "Mean BMI by income group",
    subtitle = "Adults age 20-80; unweighted descriptive summary",
    x = "Income group",
    y = "Mean BMI"
  ) +
  expand_limits(y = 0) +
  theme_minimal(base_size = 12)
```

## Limitations

- The table and figure use a prepared classroom dataset rather than raw NHANES files.
- The summaries are unweighted.
- The comparison is descriptive and does not establish causality.
- Missing BMI or income group values are excluded from the grouped summary.

## Plain-Language Summary

In this classroom dataset, mean BMI differs across income groups. The result is useful for practicing reproducible reporting and dashboard interpretation, but it should be described as a descriptive pattern rather than a causal finding.

## AI-Use and Audit Note

Document any AI assistance used to draft code, captions, or interpretation. State exactly which output you checked against the code, data dictionary, or rendered report.

## Publishing Checklist

- Report renders from source.
- Citations resolve.
- Figures and tables are generated by code.
- Public preview contains no restricted data or private links.
- Private preview access is granted to course staff when needed.
- Canvas submission includes the repository link, report path, and latest commit hash.

## Grounding Audit Prompt

Select three claims from the report. For each claim, write the supporting evidence source: code chunk, table, figure, or citation. Revise any claim that cannot be traced to evidence in the report.
