---
format:
  html:
    embed-resources: true
---

<!-- AI-EDIT(2026-06-22): REVIEW-row194 (MIT-078/MIT-080) — per-page embed-resources:true so the standalone downloaded HTML bundles the plot (narrow fix per instructor decision; verify it survives the book render) -->

# Starter Dashboard {.unnumbered}

## Where this fits

You named the three pathways on [dash01](dash01-overview-and-pathways.qmd) and applied the 1-1-1-1-1 rule on [dash02](dash02-from-eda-to-prototype.qmd). This page is the **required core artifact** for Week 8: the Quarto-rendered dashboard-style page you will adapt for Assignment 7. The Shiny exemplar discussed on [dash05](dash05-testing-privacy-and-shiny-demo.qmd) is a demonstration only — your A7 deliverable is built from the file on this page, not from the Shiny app.

## Goal

This starter shows how to turn the Week 7 descriptive summary into a dashboard-style KT product. It uses one editable control value, one visualization, one table, one interpretation, and one caution — the 1-1-1-1-1 rule made concrete.

## Audience Question

Audience: a public-health analyst preparing a short briefing on descriptive BMI patterns in the classroom NHANES equity dataset.

Question: among adults age 20-80, how does mean BMI vary across NHANES cycles for one selected income group?

## Setup

```{r}
#| label: dashboard-setup
#| message: false
#| warning: false
library(tidyverse)  # AI-EDIT(2026-06-23): tidyverse-default — consolidated core library() calls into library(tidyverse)
library(knitr)

candidate_paths <- c(
  "examples/nhanes-equity/data/nhanes_equity_v6.csv",
  "../../examples/nhanes-equity/data/nhanes_equity_v6.csv"
)

data_path <- candidate_paths[file.exists(candidate_paths)][1]
if (is.na(data_path)) {
  stop("Could not find examples/nhanes-equity/data/nhanes_equity_v6.csv")
}

nhanes <- read_csv(data_path, show_col_types = FALSE)
```

## Control

In a Shiny app, this would become a dropdown. In this static starter, change the value of `selected_income` and render again.

```{r}
#| label: dashboard-control
available_income_groups <- sort(unique(na.omit(nhanes$IncomeGroup)))
selected_income <- "Low Income (<1.3)"

if (!selected_income %in% available_income_groups) {
  selected_income <- available_income_groups[1]
}

selected_income
```

<!-- AI-EDIT(2026-06-11): MIT-081 — needs review -->
For example, change the line to `selected_income <- "Middle Income"` — the other valid value in the classroom dataset is `"High Income (>3.5)"`. Copy the value exactly, including the quotes and any parentheses.

## Summary

```{r}
#| label: dashboard-summary
dashboard_summary <- nhanes |>
  filter(
    Age >= 20,
    Age <= 80,
    IncomeGroup == selected_income,
    !is.na(BMI),
    !is.na(Cycle)
  ) |>
  group_by(Cycle) |>
  summarise(
    n = n(),
    mean_bmi = round(mean(BMI), 1),
    .groups = "drop"
  )

kable(dashboard_summary)
```

## Visualization

```{r}
#| label: dashboard-plot
ggplot(dashboard_summary, aes(x = Cycle, y = mean_bmi, group = 1)) +
  geom_line(linewidth = 0.8, color = "#2f6f73") +
  geom_point(size = 2.4, color = "#2f6f73") +
  labs(
    title = paste("Mean BMI for", selected_income),
    subtitle = "Adults age 20-80 in the prepared NHANES Health Equity classroom dataset",
    x = "NHANES cycle",
    y = "Mean BMI",
    caption = "Descriptive, unweighted classroom summary. Missing BMI, income group, cycle, and out-of-cohort ages are excluded."
  ) +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Interpretation

For the selected income group, the table and figure describe the mean BMI pattern across available NHANES cycles in the prepared classroom dataset. The dashboard should help the audience ask better questions about pattern, timing, and context rather than treat the line as a causal explanation.

## Limitation

This starter is unweighted and descriptive. It does not account for NHANES survey design, does not estimate national prevalence, and should not be used to make causal claims about income and BMI.

## Testing

Before submitting a dashboard-style product:

1. Render this page from the repository root.
2. Change `selected_income` to another value from `available_income_groups`.
3. Render again and check that the table, title, and plot all update.
4. Verify that the caption still names the dataset, exclusions, and limitation.
5. Record any warning or error in your testing log.

<!-- AI-EDIT(2026-06-22): REVIEW-row190 / MIT-078,MIT-080 — render steps come from the shared course include (D-003); kept here because this is the page holding the required A7 artifact -->
{{< include /_snippets/render-how.qmd >}}

HTML is the expected output for this dashboard — you do not need a PDF. If the render log warns about a missing TeX installation or a PDF problem, you can ignore it as long as the `.html` file was created. <!-- AI-EDIT(2026-06-22): REVIEW-row190 / MIT-080 — added cross-ref so the brief HTML/PDF reminder reads as intentional reinforcement of the fuller routine on dash05, not accidental duplication -->
[dash05](dash05-testing-privacy-and-shiny-demo.qmd) walks through the full six-step test routine for this same page.

## Privacy-Safe Publishing

- Use aggregate summaries, not row-level records.
- Avoid small-cell interpretations when `n` is low.
- State that outputs are descriptive and unweighted.
- Keep cached classroom data separate from any raw NHANES retrieval workflow.
- Do not upload row-level data to external AI tools.

## Where this goes next

[dash04](dash04-audience-and-ui-text.qmd) shows what counts as an audience-facing text change (allowed) versus an analytic change (not allowed for A7). [dash05](dash05-testing-privacy-and-shiny-demo.qmd) walks through how to test this page in Codespaces and shows the optional Shiny exemplar that your instructor will demo in class.
