# Starter Dashboard {.unnumbered}

## Goal

This starter shows how to turn the Week 7 descriptive summary into a dashboard-style KT product before building a full app. It uses one editable control value, one visualization, one table, one interpretation, and one caution.

The complete Shiny exemplar remains the main Week 8 app. This page is a lightweight planning and testing aid.

## Audience Question

Audience: a public-health analyst preparing a short briefing on descriptive BMI patterns in the classroom NHANES equity dataset.

Question: among adults age 20-80, how does mean BMI vary across NHANES cycles for one selected income group?

## Setup

```{r}
#| label: dashboard-setup
#| message: false
#| warning: false
library(readr)
library(dplyr)
library(ggplot2)
library(knitr)

candidate_paths <- c(
  "examples/nhanes-equity/data/nhanes_equity_v6.csv",
  "../../examples/nhanes-equity/data/nhanes_equity_v6.csv"
)

data_path <- candidate_paths[file.exists(candidate_paths)][1]
if (is.na(data_path)) {
  stop("Could not find examples/nhanes-equity/data/nhanes_equity_v6.csv")
}

nhanes <- read_csv(data_path, show_col_types = FALSE)
```

## Control

In a Shiny app, this would become a dropdown. In this static starter, change the value of `selected_income` and render again.

```{r}
#| label: dashboard-control
available_income_groups <- sort(unique(na.omit(nhanes$IncomeGroup)))
selected_income <- "Low Income (<1.3)"

if (!selected_income %in% available_income_groups) {
  selected_income <- available_income_groups[1]
}

selected_income
```

## Summary

```{r}
#| label: dashboard-summary
dashboard_summary <- nhanes |>
  filter(
    Age >= 20,
    Age <= 80,
    IncomeGroup == selected_income,
    !is.na(BMI),
    !is.na(Cycle)
  ) |>
  group_by(Cycle) |>
  summarise(
    n = n(),
    mean_bmi = round(mean(BMI), 1),
    .groups = "drop"
  )

kable(dashboard_summary)
```

## Visualization

```{r}
#| label: dashboard-plot
ggplot(dashboard_summary, aes(x = Cycle, y = mean_bmi, group = 1)) +
  geom_line(linewidth = 0.8, color = "#2f6f73") +
  geom_point(size = 2.4, color = "#2f6f73") +
  labs(
    title = paste("Mean BMI for", selected_income),
    subtitle = "Adults age 20-80 in the prepared NHANES Health Equity classroom dataset",
    x = "NHANES cycle",
    y = "Mean BMI",
    caption = "Descriptive, unweighted classroom summary. Missing BMI, income group, cycle, and out-of-cohort ages are excluded."
  ) +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Interpretation

For the selected income group, the table and figure describe the mean BMI pattern across available NHANES cycles in the prepared classroom dataset. The dashboard should help the audience ask better questions about pattern, timing, and context rather than treat the line as a causal explanation.

## Limitation

This starter is unweighted and descriptive. It does not account for NHANES survey design, does not estimate national prevalence, and should not be used to make causal claims about income and BMI.

## Testing

Before submitting a dashboard-style product:

1. Render this page from the repository root.
2. Change `selected_income` to another value from `available_income_groups`.
3. Render again and check that the table, title, and plot all update.
4. Verify that the caption still names the dataset, exclusions, and limitation.
5. Record any warning or error in your testing log.

## Privacy-Safe Publishing

- Use aggregate summaries, not row-level records.
- Avoid small-cell interpretations when `n` is low.
- State that outputs are descriptive and unweighted.
- Keep cached classroom data separate from any raw NHANES retrieval workflow.
- Do not upload row-level data to external AI tools.
