# Week 3: R with AI {.unnumbered}

## Overview

**Summary:** In Week 2 you learned how to use your Codespace, edit a Quarto document, and make your first commit. Now you learn how to speak R---not by memorizing syntax, but by directing an AI co-pilot and *auditing* what it writes. Your job is to be the pilot: you decide what the analysis should do, the AI drafts the code, and you verify every line before it flies.

**Learning Objectives:**

- Understand that R is a calculator that needs specific, case-sensitive instructions.
- Load packages with `library()` and explain why this is needed each session.
- Run a minimal `renv` checkpoint so package changes are visible.
- Use core constructs: variables (assignment), functions (with arguments), and the pipe (`%>%` / `|>`).
- Perform basic data transformations using `dplyr` verbs: `filter()`, `select()`, `mutate()`, `summarise()`, `group_by()`.
- Inspect a dataset on first contact: structure, dimensions, missingness.
- Identify control flow (`if/else`, loops) in AI-generated code and audit whether it is correct.
- Write and adapt a simple custom function.
- Apply an AI safety checklist to catch hallucinations, wrong column names, and unsafe operations.
- Document your analysis in Quarto by weaving code chunks with Markdown narrative.
- Commit incrementally as you work (reinforcing Week 2 Git basics).

## Connection

Week 2 made the project structure reproducible. This week you use that structure to load, inspect, summarize, and audit R code with AI support. Next week you will put those edits under version control.

## Case Study Data Analysis

{{< include ../../examples/nhanes-equity/snippets/case-study-data-analysis.qmd >}}

## In-Class Activity

- Load the CSV snapshot and inspect column names, row count, and missingness.
- Summarize one numeric variable and one grouping variable with clear `NA` handling.
- Ask an AI assistant to comment on your code, then audit whether the comments match the actual operations.
- Note one place where an AI assistant could hallucinate a variable name or overstate what the descriptive summary means.

Student output: a short Quarto note with one data inspection, one summary, and one AI-audit comment. Assignment link: [Assignment 2](../../assignments/assignment02-r-with-ai/README.md).

---

## `r1.qmd` Packages: Opening the Toolbox

### Install vs Load

- `install.packages("tidyverse")` — **buying** the toolbox. Done once per environment (already done in your course Codespace).
- `library(tidyverse)` — **opening** the toolbox. Required every time you restart R or reopen your Codespace.

```{r, eval=FALSE}
library(tidyverse)
```

::: {.callout-note}
If you see an error like `there is no package called 'tidyverse'`, the package is not installed in your Codespace. Use the Canvas support channel or office hours and include the full error message.
:::

### Minimal renv Checkpoint

`renv` helps record which R packages a project uses. You do not need to master `renv` this week, but you should recognize the three checkpoint commands that support reproducibility.

```{r, eval=FALSE}
# Check whether the project environment has changed.
renv::status()

# Record package changes when you intentionally add or update packages.
renv::snapshot()

# Restore the recorded package environment in a fresh setup.
renv::restore()
```

For Assignment 2, run `renv::status()` and write a short dependency note. Only run `renv::snapshot()` if you intentionally added a package.

---

## `r2.qmd` Core Constructs: Boxes, Toasters, and the Assembly Line

### Variables: The Labeled Box

R has no memory unless you store results in a named box using `<-`.

```{r, eval=FALSE}
patient_age <- 55
clinic_name <- "Vancouver General"
```

::: {.callout-warning}
**Case sensitivity:** `Age`, `age`, and `AGE` are three different boxes. This is one of the most common bugs in AI-generated code. Always check that column names match your data exactly.
:::

### Functions: The Toaster

A function takes **input** and returns **output**. Arguments (inside the parentheses) control its behaviour.

```{r, eval=FALSE}
mean(c(72, 85, 90), na.rm = TRUE)
```

- `c(72, 85, 90)` is the input (a vector of numbers).
- `na.rm = TRUE` tells R to ignore missing values. **Always look for this** when auditing AI code that works with health data --- missingness is the norm, not the exception.

### The Pipe: The Assembly Line

The pipe passes the result of one step into the next. Think of it as **"and then..."**

```{r, eval=FALSE}
data %>%
  filter(age > 65) %>%
  summarise(mean_bmi = mean(bmi, na.rm = TRUE))
```

Read this as: *"Take the data, and then keep only patients over 65, and then calculate the average BMI."*

::: {.callout-tip}
You will see two pipe symbols in the wild: `%>%` (from the `magrittr` / `tidyverse` packages) and `|>` (built into R 4.1+). They do the same thing for our purposes. If AI generates the other one, don't panic --- both work in your Codespace.
:::

---

## `r3.qmd` Data Import and First Contact

### Loading Data

Use relative paths so your code runs on any machine (including your grader's Codespace).

```{r, eval=FALSE}
my_data <- read_csv("data/patients.csv")
```

**How to get the path right:** Right-click the file in the VS Code Explorer → **Copy Relative Path** → paste it into your code.

**AI Prompt:**

> Read the CSV file at "data/patients.csv" and save it as `my_data`. Use the `readr` package.

### First Contact: Inspect Before You Analyze

Never start analysis without looking at your data. Run these four checks:

```{r, eval=FALSE}
# How big is it?
dim(my_data)          # rows x columns

# What are the column names and types?
glimpse(my_data)

# Quick summary statistics
summary(my_data)

# How much is missing?
colSums(is.na(my_data))
```

Open the **Environment** pane (top-right in RStudio or via the R extension) to see your variables and click a data frame to view it in a grid.

::: {.callout-warning}
**Health data reality:** Real datasets like NHANES have substantial missingness. Knowing *where* and *how much* is missing is not optional --- it shapes every downstream decision.
:::

---

## `r4.qmd` Data Transformation: The Five `dplyr` Verbs

These five verbs are the core of data wrangling in R. You will use them in nearly every assignment and in your term project.

### `filter()` — Keep rows that match a condition

```{r, eval=FALSE}
# Patients aged 65 and older
seniors <- my_data %>% filter(age >= 65)
```

### `select()` — Keep (or drop) columns

```{r, eval=FALSE}
# Keep only ID, age, and BMI
slim <- my_data %>% select(id, age, bmi)
```

### `mutate()` — Create or modify a column

```{r, eval=FALSE}
# Add a BMI category column
my_data <- my_data %>%
  mutate(bmi_category = case_when(
    bmi < 18.5 ~ "Underweight",
    bmi < 25   ~ "Normal",
    bmi < 30   ~ "Overweight",
    TRUE       ~ "Obese"
  ))
```

### `summarise()` — Collapse rows into a summary

```{r, eval=FALSE}
my_data %>% summarise(mean_age = mean(age, na.rm = TRUE))
```

### `group_by()` + `summarise()` — Summaries by group

```{r, eval=FALSE}
my_data %>%
  group_by(sex) %>%
  summarise(
    n = n(),
    mean_bmi = mean(bmi, na.rm = TRUE)
  )
```

**AI Prompt:**

> Using `dplyr`, filter my dataset to patients over 65, group by sex, and calculate the mean BMI for each group. Handle missing values.

::: {.callout-tip}
**Audit the AI output:** After running the AI-generated code, check that (a) the column names match your actual data, (b) `na.rm = TRUE` is present where needed, and (c) the row counts make sense (e.g., the filtered data should be smaller than the original).
:::

**Git checkpoint:** You have now loaded, inspected, and transformed data. Stage your `.qmd`, write a commit message describing what you did, and commit. Don't wait until the end.

---

## `r5.qmd` Control Flow: Auditing AI Decisions

You will not write many `if/else` blocks or loops by hand. But AI frequently generates them, so you need to **read** them.

### If/Else: The Fork in the Road

```{r, eval=FALSE}
if (mean_age > 50) {
  print("Older cohort")
} else {
  print("Younger cohort")
}
```

**Audit question:** Is the threshold (50) appropriate for my research question, or did the AI pick an arbitrary number?

### Loops: The Treadmill

```{r, eval=FALSE}
for (col in c("age", "bmi", "sbp")) {
  print(summary(my_data[[col]]))
}
```

**Audit questions:** Does the loop have a clear end condition? Could this be replaced with a simpler vectorized operation (e.g., `summarise(across(...))`)?

::: {.callout-note}
In R, loops are rarely the best tool. If AI gives you a loop, ask: *"Can this be done without a loop using dplyr or purrr?"* The answer is usually yes.
:::

---

## `r6.qmd` Writing a Simple Function

The syllabus asks you to write at least one custom function. A function packages a repeated task so you don't copy-paste code.

### Example: A Reusable Summary

```{r, eval=FALSE}
summarise_column <- function(data, column_name) {
  data %>%
    summarise(
      n_total   = n(),
      n_missing = sum(is.na(.data[[column_name]])),
      mean_val  = mean(.data[[column_name]], na.rm = TRUE),
      sd_val    = sd(.data[[column_name]], na.rm = TRUE)
    )
}

# Use it
summarise_column(my_data, "bmi")
summarise_column(my_data, "age")
```

### Your Task

1. Ask your AI co-pilot to write a function that takes a data frame and a column name and returns summary statistics.
2. **Audit the code:** Does it handle `NA`? Does it work on a numeric column? What happens if you pass a categorical column?
3. **Adapt it:** Change one thing (e.g., add a median, or add a `group_by` argument). This is the difference between accepting code blindly and understanding it.

**AI Prompt:**

> Write an R function called `summarise_column` that takes a data frame and a column name (as a string) and returns the count, number of missing values, mean, and standard deviation. Then show me how to call it on the "bmi" column of `my_data`.

---

## `r7.qmd` Debugging: Reading the Red Text

Errors are normal. AI makes mistakes, you make typos, and data surprises you.

### The "Fix It" Loop

1. **Read** the error message. Look for keywords: `not found`, `unexpected`, `non-numeric`.
2. **Copy** the full error text.
3. **Paste** it into your AI chat and ask:

> Here is an R error from my Codespace: [paste]. Explain what went wrong in one sentence and give corrected code.

### Common Errors and What They Mean

| Error message | Likely cause |
|---|---|
| `object 'X' not found` | Typo in variable name, or you forgot to run an earlier chunk |
| `could not find function "X"` | Forgot `library(tidyverse)` or misspelled the function |
| `non-numeric argument to binary operator` | Trying to do math on a text/factor column |
| `unexpected symbol` | Missing comma, parenthesis, or pipe |

::: {.callout-warning}
**AI debugging guardrail:** When AI suggests a fix, re-read the corrected code before running it. Does the fix address the *actual* error, or did AI rewrite the whole chunk and change your logic?
:::

---

## `r8.qmd` AI Safety: The Audit Checklist

Every time you accept AI-generated code, run through this checklist:

1. **Column names:** Are they spelled exactly as they appear in `glimpse(my_data)`?
2. **Data types:** Is the code doing math only on numeric columns?
3. **Missingness:** Is `na.rm = TRUE` present where needed?
4. **Privacy:** Does the code avoid printing individual patient rows that could be identifiable?
5. **Logic:** Does the code do what *you* intended, or what the AI *assumed* you meant?

### The Paper Trail Rule

Every time you use AI to write a code chunk, add a `#` comment above it explaining what you asked for and what you verified:

```{r, eval=FALSE}
# AI prompt: "Calculate mean BMI by sex, handling NAs"
# Verified: column names match, na.rm = TRUE present, output has 2 rows (M/F)
my_data %>%
  group_by(sex) %>%
  summarise(mean_bmi = mean(bmi, na.rm = TRUE))
```

This comment trail is your audit log. It demonstrates accountability (required by the course AI policy) and helps you debug later.

---

## `r9.qmd` Exporting Clean Data

After cleaning and transforming, save your work so future weeks can pick up where you left off.

```{r, eval=FALSE}
# Save a clean version for Week 4 and beyond
write_csv(my_data, "output/clean_data.csv")
```

::: {.callout-tip}
**Reproducibility habit:** Your Quarto document should contain *every step* from raw data to clean export. Anyone (including your grader) should be able to render it from scratch and get the same `clean_data.csv`.
:::

**Git checkpoint:** Your analysis is complete. Stage, commit with a clear message (e.g., `"Complete Week 3: load, inspect, transform, and export patient data"`), and Sync.

---

## `r10.qmd` Literate Analysis in Quarto

Your `.qmd` file is not just code --- it is a **document**. Weave narrative around your code chunks to explain what you are doing and why.

### Pattern

````markdown
## Data Inspection

We begin by examining the structure of the dataset to understand
variable types and missingness before any transformations.

```{{r, eval=FALSE}}
glimpse(my_data)
colSums(is.na(my_data))
```


The dataset contains `nrow(my_data)` observations across
`ncol(my_data)` variables. Notably, BMI is missing for
approximately 12% of records, which we handle with `na.rm = TRUE`
in downstream summaries.

This interleaving of code and prose is **literate programming** --- the methodology that underpins reproducible research in this course. Practice it now; you will need it for your term project report (Week 10) and final portfolio.

---

## `r11.qmd` Knowledge Check

1. What is the difference between `install.packages()` and `library()`?
2. Why is `na.rm = TRUE` important when working with health data?
3. What does `my_data %>% filter(age > 65) %>% summarise(n = n())` do? Read it aloud using "and then."
4. Name the five core `dplyr` verbs and what each does in one sentence.
5. You see the error `object 'BMI' not found` but your column is called `bmi`. What went wrong?
6. What is the difference between `%>%` and `|>`? Does it matter for this course?
7. Why should you write `#` comments above AI-generated code?
8. What four checks should you run immediately after loading a new dataset?

---

## `r12.qmd` Assignment: The "Mad Libs" Report

**Task:** Build a short Quarto report that loads, inspects, transforms, and summarizes a health dataset --- using your AI co-pilot for code generation and your own judgment for verification.

### Step-by-Step Workflow

1. Open `week3_assignment.qmd` in your Codespace.
2. **Load packages and data:** Use AI to write the `library()` and `read_csv()` calls. Verify the data loaded correctly with `dim()` and `glimpse()`.
3. **Inspect missingness:** Run `colSums(is.na(...))`. Add a Markdown paragraph interpreting what you see.
4. **Transform:** Use `dplyr` verbs to:
   - Filter to a meaningful subgroup (e.g., age ≥ 18).
   - Create at least one new column with `mutate()`.
   - Produce a grouped summary with `group_by()` + `summarise()`.
5. **Write a function:** Create (or adapt from AI) a reusable summary function. Call it on at least two different columns.
6. **Export:** Save your cleaned data to `output/clean_data.csv`.
7. **Document:** Every code chunk should have:
   - A `#` comment stating what you asked the AI and what you verified.
   - A Markdown paragraph below interpreting the result.
8. **Render** the document to verify it runs cleanly from top to bottom.
9. **Git:** Stage, write a descriptive commit message, Commit, and Sync.

### Deliverable

A rendered `.qmd` that a reader can follow as a narrative: what data you started with, what you did to it, what you found, and why your decisions were justified.

---

## `r13.qmd` Quick Reference

| Concept | R code | Notes |
|---|---|---|
| Load packages | `library(tidyverse)` | Every session |
| Read CSV | `read_csv("data/file.csv")` | Use relative paths |
| Assignment | `x <- 5` | The `<-` arrow stores a value |
| Inspect structure | `glimpse(df)` | Column names, types, preview |
| Dimensions | `dim(df)` | Rows × columns |
| Summary stats | `summary(df)` | Min, median, mean, max, NAs |
| Check missingness | `colSums(is.na(df))` | Count NAs per column |
| Filter rows | `filter(df, age > 65)` | Keep matching rows |
| Select columns | `select(df, id, age, bmi)` | Keep named columns |
| New column | `mutate(df, bmi_cat = ...)` | Create or modify |
| Summarize | `summarise(df, m = mean(x, na.rm = TRUE))` | Collapse to one row |
| Group + summarize | `group_by(df, sex) %>% summarise(...)` | Per-group summaries |
| Export | `write_csv(df, "output/clean.csv")` | Save for later |
| Pipe | `%>%` or `|>` | "And then..." |
